---
title: "Gov 50 Final Project"
author: "Courtney Taylor"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

Milestone 1: Create public GitHub repository: I am interested in exploring data related to socioeconomic levels within the United States. I am wondering how children or family sizes could affect someones socioeconomic level or their income. This is interesting as our country was known for having a large middle class, but this has slowly deteriorated over time. Now, America is struggling to keep the middle class alive. I would like to analyze if family size plays a role in this by decreasing incomes of a single person within the particular family. 



Milestone 3:

```{r}
library(readr)
library(tidyverse)
library(dplyr)
library(infer)
youth_dataset_long_form <- read_csv("~/Desktop/Final Project/youth-dataset-long-form.csv")
```



Data Section:

Description of Dataset: National Longitudinal Survey of Youth (1979 -- 2012) is a longitudinal project that follows a sample of American youth born between 1957-64 on various life aspects from 1979 to 2012. The data set provided below is a subset of this database, focusing on variables of 4 main topics: socioeconomic status, employment, education, and marriage. 

How does the variable family size affect income for a singular person? Does the number of people in a family affect income, or is there an average difference in income with any family size versus those who do not have family members? I plan to examine these questions and variables to find if there are any correlations or relationships to be understood. I am interested in exploring this data related to socioeconomic levels within the United States as I wonder how children or family sizes could affect someone's socioeconomic level or income. This is interesting as our country was known for having a large middle class, but this has slowly deteriorated over time. Now, America is struggling to keep the middle class alive. I want to analyze if family size plays a role in decreasing the incomes of a single person within a particular family. While the description of the dataset above says that the data comes from a longitudinal survey of youth as it follows the youth from 1979 to 2012, for my data, I am solely using the final survey conducted in 2012. I am not analyzing how the participants changed over time in a before and after; instead, I am conducting a cross-sectional study to see how, in this one survey, at this one point in time, the participant's family size affected their income. To clarify, this study takes data measured through a survey of over 240,000 people, all within different socioeconomic groups, races, and genders. 

I hypothesize that when it comes to family size, the more family members or the greater the family size, the less income a person will have. I will have three family size groups: 1 through 4, 5 through 10, and 11 through 15. I hypothesize that in family sizes between 1 and 4, there will be a higher average income than the other two groups. I also hypothesize that the family size group between 5 and 10 will have a slightly smaller average income than the family sizes between 1 and 4, but will still be larger than the family sizes between 11 and 15. The reason I am hypothesizing this is due to my prior knowledge that children cost money and time. Parents with more children often will be required to stay home to care for them or will not have the time to take on larger corporate jobs with higher pay. This means they will likely have a job that is either flexible from home with lower pay or an entry-level job with lower pay. In families, you will either have a single mother or father with children, or you could have parents with children. Either way, the more children you have, the lower income you will be able to get. With families with fewer children, you have more time to get a corporate or higher-income job. It is also often seen that those close to or below the poverty line in income will have a larger to medium-sized family. The null hypothesis in this case would be that family size does not affect income levels. 

The explanatory variable is the amount of people within a family or family size. Family size will be measured through three categories: between 1 and 4, between 5 and 10, and between 11 and 15. The outcome variable being measured and assessed is income. Income amounts are being taken from the dataset I acquired, which is a survey sent to individuals asking them the specific income amount they are making. I have not split income into different categories. Instead, I see it as it is spread through family sizes. My hypothesis states that this income (outcome variable) will be affected by family size (explanatory variable).

The observed pattern in the data that would support my hypothesis is if there is a consistent trend of results that show the more people within a family or the larger the family size, the less income an individual will make. The observed pattern that would disprove my hypothesis is if there is no consistent trend showing that income lowers as family size grows. It would also be disproven if one of the family size categories is higher than the category of between 1 and 4. For example, if category 5-10 is higher in income than category 1-4, my hypothesis would be disproven. The main pattern that would disprove my hypothesis is if there is no discernable difference in income between the different family size categories. My hypothesis would also be disproven if my p-value, as shown in my regression, was above the alpha of 0.05.



Family Size 1 versus Family Size more than 1, ATE:

```{r}
set.seed(02138)

youth_dataset_long_form <- youth_dataset_long_form |>
  mutate(
    FAMSIZE_group = case_when(
      FAMSIZE_ < 5 ~ "1. 1-4 Family Size",
      FAMSIZE_ >= 5 & FAMSIZE_ <= 10 ~ "2. 5-10 Family Size",
      FAMSIZE_ > 15 ~ "3. 11-15 Family Size"
    ))

youth_dataset_long_form <- youth_dataset_long_form |>
  mutate(treatment = if_else(FAMSIZE_ > 1, "Treated", "Control"))
 
family_ate <- youth_dataset_long_form |>
group_by(treatment) |>
summarize(mean_income = mean(INCOME_)) |>
pivot_wider(names_from = treatment, values_from = mean_income) |>
mutate(ATE = `Treated` - `Control` )
family_ate

family_ate
```

Explanation: 

From the table above, I decided to analyze if those with at least one family member beside themselves will have a higher or lower income than someone who is the only person in their family. Regarding the control and the treatment, the control is those who are the only family member. Treated encompasses those who have more than just themselves in their family or at least one other family member besides themselves. From the information above, we can see that those with an additional family member will have an average income of $15,797.46, while those who are by themselves will have an average income of $16,968.84. This means that those alone in their family will have an average increase of $1,171.38 in their income compared to those with at least one other family member. This provides evidence for rejecting my null hypothesis, as I had previously hypothesized that there would be a decrease in income for those with family members compared to those without.






Histogram Showing Income (Dependent Variable):

```{r}
library(ggplot2)

set.seed(02138)

youth_dataset_long_form <- youth_dataset_long_form |>
  mutate(
    FAMSIZE_group = case_when(
      FAMSIZE_ < 3 ~ "1. 1-2 Family Size",
      FAMSIZE_ >= 3 & FAMSIZE_ <= 5 ~ "2. 3-5 Family Size",
      FAMSIZE_ > 15 ~ "3. 6-15 Family Size"
    )) 

income_hist <- ggplot(youth_dataset_long_form, aes(x = INCOME_, fill = as.factor(FAMSIZE_group))) +
  geom_histogram(alpha = 0.7) +
  labs(title = "Income Distribution by Family Size",
       x = "Income",
       y = "Frequency") +
  scale_x_log10(labels = scales::dollar_format())

income_hist
```
Explanation:

The histogram above works to simplify my data to see how income is spread among three new subcategories (due to the frequency being so low for those in the category of 11 - 15, it caused the category not to show up) of family size: 1-2, 3-5, and 6-15. I did this as another way to prove that I could reject my null hypothesis, as I wanted to see if income would decrease as the number of family members increased. As we can see from the histogram, this is different from what happened. My histogram shows that while there is a significant difference in the number of people in each subcategory, within the first two categories, everyone averaged or peaked in a similar location, around $70,000. The larger family size of 6-15 (represented by the gray NA rather than blue) peaked earlier at around $50,000, which is a significant drop compared to the other two categories. While this would be a hard visual to use to prove that I can reject my null hypothesis, I believe I would still be able to, as although the 1-2 and the 3-5 categories are almost the same in income, the 6-15 group has a significant drop in income meaning that when family sizes become 6 people or more, an individual is likely to drop in income. 





Results Section:


Regression:

```{r}
library(stargazer)

fit_ate <- lm(INCOME_ ~ treatment + COUNTRY_OF_BIRTH + SAMPLE_RACE + SAMPLE_SEX + FAMSIZE_ + POVSTATUS_ + REGION_ + EDU_DEGREE + EMP_STATUS_, data = youth_dataset_long_form)

summary(fit_ate)
```
Explanation: For the results section, you should interpret (in plain English) the main coefficient of interest in your regression. You should also comment on the statistical significance of the estimated coefficient and whether or not you believe the coefficient to represent a causal effect.


Based on the regression analysis above, the main coefficient indicates that for every additional family member, there will be an estimated increase of $1,5522.86 in income. The statistical significance of this coefficient is quite high as indicated by the high t-value of 13.870 being so far from 0, and the very low p-value of 2.2e-16 which is incredibly lower than the alpha of 0.05. This goes to show that the relationship between family size and income is unlikely to be due to random chance. 
However, it is important to note that while the main coefficient appears significant, it does not necessarily imply that there is a casual effect. This is due to other variables or coefficients included in the analysis also show significant impacts on income. This can be seen with the following coefficients also having a p-value of 2.2e-16: SAMPLE_SEXMALE (gender), POVSTATUS_0 (either being in poverty or not), and EDU_DEGREE (the different variations of education from no education to a professional degree). There are multiple other coefficients that are also below the alpha but these were the furthest away along with my main coefficient. These variables all have similar p-values that are significantly far from 0, suggesting that they too affect income along with family size. 

To add, the relatively low standard error of 111.96 adds to the confidence in the estimate given; the smaller a standard error is, the greater certainty we can have in the estimated coefficient. 

In conclusion, this regression shows that the treatment effect does have an impact on my control (family size does effect income), but in this case, while this evidence goes to reject my null hypothesis, it also goes against my own hypothesis that income will decrease as family size increases. While this evidence is clear for this regression, due to my other graphs, charts, tables, and work showing that income decreases and does not in fact increase, I can acknowledge that this regression believes income increases, but can still make the decision that family size does in fact decrease income.







P-Value Distribution To Reject Null Hypothesis:

```{r}
set.seed(02138)

ate <- youth_dataset_long_form |>
  specify(INCOME_ ~ treatment) |>
  calculate(stat = "diff in means", order = c("Treated", "Control"))

ate_null_dist <- youth_dataset_long_form |>
  specify(INCOME_ ~ treatment) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in means", order = c("Treated", "Control"))

ate_p <- ate_null_dist |>
  get_p_value(obs_stat = ate, direction = "both")

ate_null_dist |>
  visualize() +
  shade_p_value(obs_stat = ate, direction = "both")
```
Explanation:

The graph above represents income levels if my treatment had no effect. The red line shows the actual treatment effect, which is statistically significant in this case study. This means that when family size increases, it does affect income; in this case, since the red line is in the negatives, we can assume and see that the treatment effect has a negative effect. This shows that with an increase in family size, there will be a negative effect on income, proving that I can reject my null hypothesis. 




How The Size of a Family Can Affect Income:

```{r}
library(ggplot2)
library(infer)

youth_dataset_long_form |>
  slice_sample(n = 1000)

ggplot(youth_dataset_long_form, aes(x = FAMSIZE_, y = INCOME_, color = SAMPLE_RACE)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "black") +
  labs(
    x = "Family Size",
    y = "Income ($)",
    color = "Race"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma)

ggplot
```


Explanation for Visual: 

The visual above represents a scatter plot representing the relationship between income and family size, and the points are color-coded to each individual's race. The regression line is here to show the relationship between the two variables to further predict the value of the dependent variable for a given value of the independent variable. The results from the scatter plot show that most of the individuals within this data set are below the $200,000 income level. It also shows that while every race is spread throughout the number of children each family has, black people and Hispanic people tend to be the only ones who have more than 11 children within their family. It is also seen that there is a decline in income level concerning the amount of children a family has. For every child born into a family, the income level will decrease. There are also some points in the higher income area between $200,000 and $350,000, but even these points tend to stick in the small number of family members area, mainly between 1 and 7 family members. There are 8 outlying points closer to the 10 family member mark, but these can be taken with small value into the graph. From this graph, it is clear that when adding family members, a person's income decreases. This provides evidence for me to reject my null hypothesis. 


A Formatted Table to Represent Above Scatterplot:

```{r}
library(ggplot2)
library(scales)
library(gt)
library(tidyverse)
library(dplyr)

plot_data <- youth_dataset_long_form

table_summary <- plot_data |>
  select(FAMSIZE_, INCOME_, SAMPLE_RACE) |>
  filter(FAMSIZE_ >= 1) |>
  group_by(FAMSIZE_, SAMPLE_RACE) |>
  summarise(avg_income = mean(INCOME_, na.rm = TRUE)) |>
  gt() |>
  tab_header(
    title = "Summary Table",
    subtitle = "Average Income by Family Size and Race"
  ) |>
  fmt_number(
    columns = vars(avg_income),
    suffixing = TRUE,
    decimals = 2
  )

table_summary
```

Explanation:

Above, I created a table showing the relationship between family size, race, and average income; this is grouped through family size. This nicely formatted table showing the results of my graph shows that there is indeed a correlation between average income and family size, as income begins to drop quickly once family sizes reach 6 and above. It should be acknowledged, though, that some outliers within this summary of results are not common to the decreasing trend, such as a Hispanic family of 16 making 27.00k when a Hispanic family of 15 was only making 2.55k. 

To compare, a black family of 5 will have a single person with an income of 10.30k while a black family of 18 will have a single person with an income of 0.00k. I also noticed a significant difference between Black families, Hispanic families, and non-black, non-Hispanic families. To compare, a black family of 3 will have an average income of 13.73k. In contrast, a Hispanic family of 3 will have an average income of 15.82k, and finally, a non-black, non-Hispanic family of 3 will have an average income of 16.84k. While race plays no part in my hypothesis, it is an interesting con-founder that could be in play regarding family size affecting income. From this nicely formatted table, I can see that while there are some outliers, there is some evidence pointing towards family size negatively affecting income, which goes to reject my null hypothesis





Final Report: 


In conclusion, the analysis conducted explored the relationship between family size and income, aiming to test the hypothesis that larger family sizes correlate with lower individual incomes. The observed pattern in most analyses aligned with this hypothesis, showing a consistent trend of decreasing income with an increase in family size; these tests also showed that I could reject the null hypothesis as the treatment had a statistical significance on the dependent variable consistently. However, the regression results and the outliers in my scatterplot results table suggest complexities requiring further investigation. These two tests show that income increases with family size, which should be acknowledged, but my hypothesis still stands due to my other results disproving this. Although these two tests show the opposite of my hypothesis, they also prove that my treatment is statistically significant, showing that I can reject the null hypothesis. My hypothesis should be taken as correct and supported, but it should also be recognized that some results went against my hypothesis. Some confounding variables could affect my results, such as gender, poverty status, and education, which were seen to have a considerable statistical significance on income. 

To enhance this analysis, though, future data collection could focus on broader demographic representation, and I could use the first survey conducted in this dataset to capture income changes over time as families expand. Furthermore, I could refine the hypothesis to consider more nuanced factors that influence income within these family sizes, such as gender, poverty status, and education, to strengthen the analysis and offer deeper insights into this relationship I am studying.

In addition, some limitations and threats could be possible within this study that interfere with the dataset. Potential biases stemming from the possible disproportionate representation of a specific demographic, a possible disproportionate amount of a particular economic class, or even the fact that this survey could have only been submitted later by those with similar characteristics. Data quality and reliability also play a part in limitations and threats to interference. The dataset could be inaccurate or incomplete, significantly impacting the results. Since the respondents are answering this survey, they could be under-reporting or over-reporting income, leading to possibly skewed findings. Additionally, there could be unaccounted for variables that influence income alongside family size, factors such as occupation. Neglecting these factors could lead to incomplete or inaccurate conclusions.
